---
title: "Class 09: Halloween Mini-Project"
author: "Safiya Sayd (PID: A18027139)"
format: pdf
toc: TRUE
---

## 1. Importing candy data

```{r}
candy_file <- "https://raw.githubusercontent.com/fivethirtyeight/data/master/candy-power-ranking/candy-data.csv"
candy = read.csv(candy_file, row.names=1)
head(candy)
```

> Q1. How many different candy types are in this dataset?

```{r}
nrow(candy)
```

> Q2. How many fruity candy types are in the dataset?

```{r}
sum(candy$fruity)
```

##2. What is your favorate candy?

```{r}
candy["Twix", ]$winpercent
```

> Q3. What is your favorite candy in the dataset and what is it’s winpercent value?

My favorite candy is Reese's pieces
```{r}
candy["Reese's pieces", ]$winpercent
```

> Q4. What is the winpercent value for “Kit Kat”?

```{r}
candy["Kit Kat", ]$winpercent
```

> Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy["Tootsie Roll Snack Bars",]$winpercent
```

"skimr" package gives you a quick overview of dataset
```{r}
library("skimr")
skim(candy)
```

> Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

winpercent has a different scale compared to the other variables with a scale of 0-100 while the rest are on a scale of 0-1.

> Q7. What do you think a zero and one represent for the candy$chocolate column?

```{r}
candy$chocolate
```

I think it represents if the candy is or isn't chocolate based. It's assigned a value of 1 if its a chocolate based candy and 0 if it isn't. 

> Q8. Plot a histogram of winpercent values

```{r}
library(ggplot2)
ggplot(candy) + 
  aes(winpercent) +
  geom_histogram(binwidth = 10)
```

> Q9. Is the distribution of winpercent values symmetrical?

The distribution is not symmetrical, there is a longer tail on the right side. Distribution is skewed to right. 

> Q10. Is the center of the distribution above or below 50%?

To find whether the center of distribution is above or below 50% you can use the median. If the value is less that 50 percent the center is below 50 and if its above 50% the center is above 50 percent. 

```{r}
median(candy$winpercent)
```

The center of distribution is below 50%. 

> Q11. On average is chocolate candy higher or lower ranked than fruit candy?

```{r}
mean(candy$winpercent[candy$chocolate==1])
mean(candy$winpercent[candy$fruity==1])
```
Chocolate is higher ranked then fruity 

> Q12. Is this difference statistically significant?

t test helps us compare the average win percentage of chocolate compared to fruity

```{r}
t.test(winpercent ~ chocolate, data = candy, var.equal = FALSE)
```

p-value tells you whether the difference is seen is real or by chance. When the p value is greater the 0.05 it safe to say that it is statistically significant but if its equal to or less then 0.05 it isn't significant and considered random!

To test the reliability of the t-test it is important to see if the win percentage is normally distributed using the normality test. 
```{r}
shapiro.test(candy$winpercent[candy$chocolate ==1])
shapiro.test(candy$winpercent[candy$fruity ==1])
```

p-value is more the 0.05, the data is normal! This means the t-test can be trusted. It is safe to assume that the difference is not statistically significant. 

##3. Overall Candy Rankings

> Q13. What are the five least liked candy types in this set?

```{r}
head(candy[order(candy$winpercent),], n =5)
```

> Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
head(candy[order(-candy$winpercent),], n =5)
```
```{r}
library(dplyr)
candy %>% arrange(winpercent) %>% head(5)
```
```{r}
library(dplyr)
candy %>% arrange(desc(winpercent)) %>% head(5)
```
I prefer using dplyr because its cleaner and easier to read.   

> Q15. Make a first barplot of candy ranking based on winpercent values.

```{r}
library(ggplot2)

ggplot(candy) + 
  aes(winpercent, rownames(candy)) +
  geom_col()
```

> Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy), winpercent) ) +
  geom_col()
```
Customizing a color scheme by candy type
```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"
```

```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols) 
```

> Q17. What is the worst ranked chocolate candy?

Sixlets 

> Q18. What is the best ranked fruity candy?

Starbursts

##4. Taking a look at pricepercent

```{r}
library(ggrepel)

# How about a plot of price vs win
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 5)
```

> Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

Chocolate candies, specifically Reese's miniture

> Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```

##5 Exploring the correlation structure

seeing how variables interact with one another
```{r}
library(corrplot)
```

```{r}
cij <- cor(candy)
corrplot(cij)
```

> Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

Chocolate and fruity

> Q23. Similarly, what two variables are most positively correlated?

Chocolate and bar

##6. Principal Component Analysis

```{r}
pca <- prcomp(candy, scale=FALSE)
summary(pca)

```


```{r}
plot(pca$x[,1], pca$x[,2], col = my_cols, pch=16)
```

```{r}
my_data <- cbind(candy, pca$x[,1:3])
```

```{r}
p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)

p
```

```{r}
library(ggrepel)

p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```

```{r}
library(plotly)
```

```{r}
ggplotly(p)
```

```{r}
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```

> Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

Fruity, Hard, and pluribus are strongly by PC1 in positive direction. This makes sense to me.