---
title: "Class12 Pt.2 (Population analysis) [Q13 Q14 BoxPlot]"
author: "Safiya Sayd (PID: A18027139)"
format: html
date: 02/13/2025
---

##Section 4: Population Scale Analysis

> Q13: Read this file into R and determine the sample size for each genotype and their
corresponding median expression levels for each of these genotypes.

How many samples do we have?
```{r}
expr <- read.table("rs8067378_ENSG00000172057.6.txt")
head(expr)
```

```{r}
nrow(expr)
```
Sample size for each genotype
```{r}
table(expr$geno)
```
Median expression levels for each of these genotypes
```{r}
tapply(expr$exp, expr$geno, median)
```

> Q14: Generate a boxplot with a box per genotype, what could you infer from the relative
expression value between A/A and G/G displayed in this plot? Does the SNP effect the
expression of ORMDL3? 

```{r}
library(ggplot2)
```


Make a boxplot
```{r}
ggplot(expr, aes(x=geno, y=exp, fill = geno)) +
  geom_boxplot(notch=TRUE) +
  geom_jitter(shape = 16, position = position_jitter(0.2), alpha = 0.5, color = "black")+
  labs(title = "Expression Levels by Genotype",
       x = "Genotype",
       y = "Expression") + 
  theme_minimal()
```
G allele is associated with reduced expression of gene. This can be interpreted as SNP affecting ORMDL3 expression. 